---
title: "transit cities"
author: "Zhanchao Yang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidycensus)
library(tidyverse)
library(readxl)
```
##Data prepare
### Data input
```{r, message=FALSE,warning=FALSE}
FARS <- read.csv("data/mastersheetFARS.csv")
cbsa <- read_excel("data/cbsa_list.xlsx")
```

### Re-create the FIPS code and filter the data
filter to the year 2012-2022 and only include the important code
```{r}
FARS<-FARS %>%
  filter(YEAR>2011)%>%
  select(YEAR,STATE,COUNTY,fatality_count)
```

FIPS code creation
```{r}
FARS$FIPS<-sprintf("%02d%03d", FARS$STATE, FARS$COUNTY)
```

##CBSA
cbsa code filtering only include selecting counties
cities:  NYC, Chicago, Philadelphia, LA, D.C., Boston, Seattle, SF, Baltimore, Pittsburgh
cbsa code for 10 cities on the list 
`msa_codes <- c(35620, 16980, 37980, 31080, 47900, 14460, 42660, 41860, 12580, 38300)`

```{r}
MSA<- c(35620, 16980, 37980, 31080, 47900, 14460, 42660, 41860, 12580, 38300)
cbsa_10<-cbsa%>%
  filter(`CBSA Code` %in% MSA)%>%
  rename(cbsa_code= `CBSA Code`, cbsa_name= `CBSA Title`)

cbsa_10<- cbsa_10%>%
  select(cbsa_code,cbsa_name,`FIPS State Code`, `FIPS County Code`)%>%
  rename(STATE=`FIPS State Code`, COUNTY=`FIPS County Code`)
```


```{r}
cbsa_10$COUNTY<-as.numeric(cbsa_10$COUNTY)
cbsa_10$STATE<-as.numeric(cbsa_10$STATE)
cbsa_10$FIPS<-sprintf("%02d%03d", cbsa_10$STATE, cbsa_10$COUNTY)
```

Filtered out FARS Data
```{r}
FARS<-FARS%>%
  filter(FIPS %in% cbsa_10$FIPS)
```

### Aggregating the FARS by years and county
```{r, message=FALSE,warning=FALSE}
FARS_final<-FARS%>%
  group_by(YEAR,FIPS)%>%
  summarise(fatality_count=sum(fatality_count))
```

### join the county data into the CBSA

```{r}
cbsa_10_join <- left_join(cbsa_10, FARS_final, by = "FIPS")
```

aggregate the data to cbsa
```{r, message=FALSE,warning=FALSE}
cbsa_10_summarize<-cbsa_10_join%>%
  group_by(YEAR,cbsa_code,cbsa_name)%>%
  summarise(fatality_count=sum(fatality_count))
```


```{r}
cbsa_10_summarize<-cbsa_10_summarize%>%
  filter(!is.na(fatality_count))
```

### write a function to get the ACS by year
```{r, message=FALSE,warning=FALSE}
get_MSA<- function(year){
  temp <- get_acs(
    geography = "metropolitan statistical area/micropolitan statistical area",
    variables = "B01003_001",
    year = year,
    survey = "acs1"
  ) %>%
    filter(GEOID %in% MSA) %>%
  return(temp)
}
```

### pull the MSA census data by year
```{r, message=FALSE,warning=FALSE}
MSA_2012<-get_MSA(2012)%>%
  mutate(YEAR=2012)
MSA_2013<-get_MSA(2013)%>%
  mutate(YEAR=2013)
MSA_2014<-get_MSA(2014)%>%
  mutate(YEAR=2014)
MSA_2015<-get_MSA(2015)%>%
  mutate(YEAR=2015)
MSA_2016<-get_MSA(2016)%>%
  mutate(YEAR=2016)
MSA_2017<-get_MSA(2017)%>%
  mutate(YEAR=2017)
MSA_2018<-get_MSA(2018)%>%
  mutate(YEAR=2018)
MSA_2019<-get_MSA(2019)%>%
  mutate(YEAR=2019)
MSA_2021<-get_MSA(2021)%>%
  mutate(YEAR=2021)
MSA_2022<-get_MSA(2022)%>%
  mutate(YEAR=2022)

MSA_ACS<-rbind(MSA_2012,MSA_2013,MSA_2014,MSA_2015,MSA_2016,MSA_2017,MSA_2018,MSA_2019,MSA_2021,MSA_2022)
```

**Note: 2012 LA Metropolitan area is redistrict, so that we cannot get the ACS data directly from ACS, my suggestion is drop 2012 for LA MSA**

### pull up the 2020 census data
```{r, message=FALSE,warning=FALSE}
MSA_2020<-get_decennial(
  geography = "metropolitan statistical area/micropolitan statistical area",
  variables = "P1_001N",
  year = 2020,
) %>%
  filter(GEOID %in% MSA) %>%
  mutate(YEAR=2020)
```
### join the census and ACS data into the FARS 
```{r}
census_cbsa<- left_join(cbsa_10_summarize, MSA_ACS, by = c("cbsa_code" = "GEOID", "YEAR" = "YEAR")) %>%
  left_join(MSA_2020, by = c("cbsa_code" = "GEOID", "YEAR" = "YEAR"))

census_cbsa<-census_cbsa%>%
  select(YEAR,cbsa_code,cbsa_name,fatality_count,value,estimate)%>%
  mutate(population=ifelse(is.na(value),estimate,value))
```

### calculate the fatality rate
```{r}
census_cbsa<-census_cbsa%>%
  mutate(fatality_rate=fatality_count/population*100000)
```

```{r}
write.csv(census_cbsa,"data/MSA_transit_cities_fatality_rate.csv")
```




## Top 10 Transit Cities (Counties level)

### filter to only 10 cities respective counties

New York City, NY (covers multiple counties):
- Bronx County: 36005
- Kings County (Brooklyn): 36047
- New York County (Manhattan): 36061
- Queens County: 36081
- Richmond County (Staten Island): 36085

**Question: Should we include the all five counties in the NYC?**

Chicago, IL (Cook County): 17031

Philadelphia, PA (Philadelphia County): 42101

Los Angeles, CA (Los Angeles County): 06037

Washington, D.C. (District of Columbia): 11001

Boston, MA (Suffolk County): 25025

Seattle, WA (King County): 53033

San Francisco, CA (San Francisco County): 06075

Baltimore City (independent): 24510

Pittsburgh, PA (Allegheny County): 42003
```{r}
county<- c("36005","36047","36061","36081","36085","17031","42101","06037","11001","25025","53033","06075","24510","42003")

FARS_county<-FARS_final%>%
  filter(FIPS %in% county)
```

### Gathering ACS population data
function for ACS
```{r}
get_county<- function(year){
  temp <- get_acs(
    geography = "county",
    variables = "B01003_001",
    year = year,
    survey = "acs1",
  ) %>%
    filter(GEOID %in% county)%>%
    mutate(YEAR=year)%>%
  return(temp)
}
```

pull out ACS
```{r, message=FALSE,warning=FALSE}
county_2012<-get_county(2012)
county_2013<-get_county(2013)
county_2014<-get_county(2014)
county_2015<-get_county(2015)
county_2016<-get_county(2016)
county_2017<-get_county(2017)
county_2018<-get_county(2018)
county_2019<-get_county(2019)
county_2021<-get_county(2021)
county_2022<-get_county(2022)
county_sum<-rbind(county_2012,county_2013,county_2014,county_2015,county_2016,county_2017,county_2018,county_2019,county_2021,county_2022)
```


